Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-10 paper/paper_draft.qmd
Update author order, affiliations and contributions

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great 
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines 
    in coral cover from thermal stress events. Understanding drivers behind reef resilience 
    to climate impacts is critical for conservation planning and intervention strategies. A 
    scenario-based population modelling approach was adopted with larval connectivity dynamics, 
    thermal adaptation, and environmental factors to assess coral reef resilience across GBR 
    locations under future climate projections. Using five Global Climate Models (GCMs) 
    spanning five Equilibrium Climate Sensitivities,  reefs were analysed for coral cover 
    time series and their ability of reefs to maintain positive carbonate production budgets 
    under future conditions. Contrary to expectations, larval connectivity patterns did not 
    correlate with increases in maintenance of positive carbonate budgets. Instead, reef depth 
    emerged as the primary predictor of cover-time series resilience and maintenance of positive 
    carbonate budgets, with deeper reefs benefiting from thermal refugia that reduce bleaching 
    impacts. These findings suggest that depth provides critical 
    protection against thermal stress, contrasting current intervention practices that primarily target easily 
    accessible sites in the 1-10m depth range. These results have important implications 
    for reef management, indicating that depth refugia should be prioritized for further 
    consideration in conservation planning alongside connectivity modelling advancements 
    to maximize coral survival under continuing climate change. 
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

The full dataset consisted of the time series cluster assignment for each of the 2197 reefs
under the each GCM (totalling 10,985 samples). The dataset was randomly split into training
(60%) and testing (40%) subsets to train and assess classification performance. The RF
classifier achieved 80% accuracy on the training set and 60% on the test set.

A 3-fold cross-validation was also conducted, arriving at a similar 59% accuracy rate. F1
performance scores for Low, Medium, and High clusters were determined to be 0.70, 0.43 and
0.64 respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with the test set
(consisting of 4394 samples) (@stbl-confusion-matrix; @stbl-rf-cluster-performance). Lower
performance with the Medium cluster reflects the inherent difficulty of classifying
transitional states in natural systems. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier, while not
accurate, is still sufficiently informative for the purpose of assessing the influence of
reef characteristics.

@@

**Benjamin Grier:** conceptualisation, data curation, formal analysis, investigation,
methodology, visualization, writing – original draft, writing – review and editing. **Takuya
Iwanaga:** conceptualisation, data curation, formal analysis, investigation, methodology,
visualization, writing – original draft, writing – review and editing. **\[BG: To be
completed.\] ...**

@@

A global sensitivity analysis was performed using Shapley Effects **goda etal** implemented
in the SAShE.jl package **SAShE citation**. For this analysis, 10,000 samples were simulated
for a single reef and timestep. Two metrics were considered, difference in coral cover and
difference in evenness. The factors analysed were heritability, mean colony diameter, linear
extension, mortality base rate, mean and standard deviation of the tolerance distribution,
linear extension scale factor, mortality rate scale factor reef habitable area, reef depth,
DHW, and initial coral cover. @sfig-coralblox-sa shows the result of such analysis for
different ranges of DHW. In both cases a similar effect takes place: when DHW = 0, initial
cover is the most influential factor then, when DHW \> 0, depth and DHW start to appear as
influential factors and, as DHW values get higher, depth becomes more influential in
detriment of DHW.

@@

The CID time series clustering metric corrects the Euclidean distance between pairs of time
series with their relative complexity (which is the quotient of their variability). This is
to ensure reefs are clustered based on their underlying temporal dynamics rather than
aggregating the coral cover across all timesteps. The process creates a n \* n distance
matrix (where n is the number of time series in the group) that is used as an input to a
k-medoids clustering algorithm [cf. @steinmann2020]. The three reef clusters are labelled
with qualitative-ordered labels from Low, Medium and High based on their median cover
relative to each other cluster over the 2030-2060 period. The selected time frame prevents
the classification from being overly influenced by any steep declines in the initial years
of simulations and avoids the post-2060 time period where reefs are projected to be under
severe stress and in a declined state [@bozec2025].

@@

![Bar plot indicating the Spearman's rank correlation values for the number of years above
carbonate budget threshold and, median reef depth and log10 weighted incoming connectivity
values. Depth correlation values are displayed in the left plot, while connectivity
correlation values are displayed in the right plot. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_correlations.png){#sfig-access-esm1-5-carb-budget-corr
fig-align="center" width="14.82cm"}

>>>>

---
title: Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Pedro Ribeiro De Almeida
    email: p.ribeirodealmeida@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Chinenye Ani
    email: c.ani@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
  - name: Sam Matthews
    email: s.matthews@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Townsville, Queensland, Australia
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great 
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines 
    in coral cover from thermal stress events. Understanding drivers behind reef resilience 
    to climate impacts is critical for conservation planning and intervention strategies. A 
    scenario-based population modelling approach was adopted with larval connectivity dynamics, 
    thermal adaptation, and environmental factors to assess coral reef resilience across GBR 
    locations under future climate projections. Using five Global Climate Models (GCMs) 
    spanning five Equilibrium Climate Sensitivities,  reefs were analysed for coral cover 
    time series and their ability of reefs to maintain positive carbonate production budgets 
    under future conditions. Contrary to expectations, larval connectivity patterns did not 
    correlate with increases in maintenance of positive carbonate budgets. Instead, reef depth 
    emerged as the primary predictor of cover-time series resilience and maintenance of positive 
    carbonate budgets, with deeper reefs benefiting from thermal refugia that reduce bleaching 
    impacts. These findings suggest that depth provides critical 
    protection against thermal stress, contrasting current intervention practices that primarily target easily 
    accessible sites in the 1-10m depth range. These results have important implications 
    for reef management, indicating that depth refugia should be prioritized for further 
    consideration in conservation planning alongside connectivity modelling advancements 
    to maximize coral survival under continuing climate change. 
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

The full dataset consisted of the time series cluster assignment for each of the 2197 reefs
under the each GCM (totalling 10,985 samples). The dataset was randomly split into training
(60%) and testing (40%) subsets to train and assess classification performance. The RF
classifier achieved 80% accuracy on the training set and 60% on the test set. A 3-fold
cross-validation was also conducted, arriving at a similar 59% accuracy rate. F1 performance
scores for Low, Medium, and High clusters were determined to be 0.70, 0.43 and 0.64
respectively, with precision of 0.62, 0.51 and 0.66 when evaluated with the test set
(consisting of 4394 samples) (@stbl-confusion-matrix; @stbl-rf-cluster-performance). Lower
performance with the Medium cluster reflects the inherent difficulty of classifying
transitional states in natural systems. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier, while not
accurate, is still sufficiently informative for the purpose of assessing the influence of
reef characteristics.

@@

**Benjamin Grier:** conceptualisation, data curation, formal analysis, investigation,
methodology, visualization, writing – original draft, writing – review and editing. **Takuya
Iwanaga:** conceptualisation, data curation, formal analysis, investigation, methodology,
visualization, writing – original draft, writing – review and editing. **Pedro Ribeiro de
Almeida:** data curation, model development, analysis, writing - methodology, writing -
review and editing, visualisation. **Chinenye Ani:** data curation, writing - methodology,
writing - review and editing. **Sam Matthews:** Writing - review and editing.

@@

A global sensitivity analysis was performed using Shapley Effects [@goda2021] implemented in
the SAShE.jl package [@ribeirodealmeida2025a]. For this analysis, 10,000 samples were
simulated for a single reef and timestep. Two metrics were considered, difference in coral
cover and difference in evenness. The factors analysed were heritability, mean colony
diameter, linear extension, mortality base rate, mean and standard deviation of the
tolerance distribution, linear extension scale factor, mortality rate scale factor reef
habitable area, reef depth, DHW, and initial coral cover. @sfig-coralblox-sa shows the
result of such analysis for different ranges of DHW. In both cases a similar effect takes
place: when DHW = 0, initial cover is the most influential factor then, when DHW \> 0, depth
and DHW start to appear as influential factors and, as DHW values get higher, depth becomes
more influential in detriment of DHW.

@@

The CID time series clustering metric corrects the Euclidean distance between pairs of time
series with their relative complexity (which is the quotient of their variability). This is
to ensure reefs are clustered based on their underlying temporal dynamics rather than
aggregating the coral cover across all timesteps. The process creates a $n * n$ distance
matrix (where $n$ is the number of time series in the group) that is used as an input to a
k-medoids clustering algorithm [cf. @steinmann2020]. The three reef clusters are labelled
with qualitative-ordered labels from Low, Medium and High based on their median cover
relative to each other cluster over the 2030-2060 period. The selected time frame prevents
the classification from being overly influenced by any steep declines in the initial years
of simulations and avoids the post-2060 time period where reefs are projected to be under
severe stress and in a declined state [@bozec2025].

@@

![Bar plot indicating the Spearman's rank correlation values for the number of years above
carbonate budget threshold ($theta$) and, median reef depth and log10 weighted incoming
connectivity values. Depth correlation values are displayed in the left plot, while
connectivity correlation values are displayed in the right plot. Displays results for
ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_correlations.png){#sfig-access-esm1-5-carb-budget-corr
fig-align="center" width="14.82cm"}
